A Pitman-Yor process prior for estimating shared evolutionary events

### 2021-03-03 manuscript/ms.tex
Working on manuscript.
Adding methods subsection about how we quantify model performance.
Incorporating better plots from main project repo and relegating
other plots to the supporting info.

\author[1]{Tashitso Anamza}
\author[1]{Jamie R.\ Oaks \thanks{Corresponding author: \href{mailto:joaks@auburn.edu}{\tt joaks@auburn.edu}}}
\affil[1]{Department of Biological Sciences \& Museum of Natural History, Auburn University, Auburn, Alabama 36849}

@@

\subsubsection{An empirical application}

@@

>>>>

\author[1]{Tashitso Anamza}
\author[1]{Tanner Myers}
\author[1]{Randy Klabacka}
\author[1]{Perry L.\ Wood, Jr.}
\author[1]{Claire Tracy}
\author[1]{Kerry Cobb}
\author[1]{Matthew Buehler}
\author[1]{Jamie R.\ Oaks \thanks{Corresponding author: \href{mailto:joaks@auburn.edu}{\tt joaks@auburn.edu}}}
\affil[1]{Department of Biological Sciences \& Museum of Natural History, Auburn University, Auburn, Alabama 36849}

@@

\subsubsection{Quantifying the performance on simulated data}

The primary goal of methods like \ecoevolity is to infer the number of
divergence events and the assignment of the comparisons to those events
(\etimesets).
To quantify the error in estimating \etimesets from each simulated \dataset, we
used the partition distance between the true \etimesets and those sampled from
the posterior distribution of the DP, PYP, and \wunif models
\citep{Regnier1983,Gusfield2002}.
In our context, the partition distance between the true and sampled \etimesets
is the minimum number of divergence comparisons that need to be removed for the
assignment of the remaining comparisons to divergence events to match between
the two \etimesets.
% Below we will refer to this distance as \etimesets error.
To calculate this distance we used the Hungarian (or Kuhn-Munkres) algorithm
\citep{Kuhn1955,Munkres1957}
as implemented in Version 1.1.4 of the Munkres Python package
\citep{Clapper2020}.

In addition to partition distances as metric for \etimesets error, we used
three additional statistics to quantify performance in estimating \etimesets:
$p(\hat{\etimesets} =\etimesets)$---
    the proportion of simulation replicates in which the true \etimesets had
    the maximum posterior probability,
% However, this metric can be misleading.
% For example, in our simultaneous-divergence and independent-divergence simulations,
% the true divergence model has the maximum prior probability simply because there
% is only a single model with one and 10 divergences.
% Thus, if the data provide no information, there is a 50\% chance of the true
% model having the maximum posterior probability.
$p(\etimesets \given \alldata)$---
    the posterior probability of the true divergence model,
and
$p(\etimesets \in \textrm{95\% CS})$---
    the proportion of simulation replicates for which he true \etimesets is
    included in the 95\% posterior credibility set of divergence models (i.e.,
    coverage).

\subsection{An empirical application}

@@

When we compare the error in estimating the divergence model (\etimesets) and
the posterior support for the true \etimesets, we see that the WU model
underperforms relative to the DP and PYP models when applied to data generated
under the simultaneous, DP, and PYP models
(first two rows of \fig~\ref{fig:modelperformancegrid}).
The DP and PYP models show more robustness when applied to
data generated under the WU model.
As expected, we find the simulations where all 10 pairs diverged independently
to be the most challenging scenario for all three models.
The PYP model performs better than the DP and \wunif model
when analyzing these most challenging \datasets
(right-most column of \fig~\ref{fig:modelperformancegrid}).
This is true across all of the \dataset sizes we simulated
under the independent-divergences model
(\fig~\ref{fig:modelperformancebysize})
However, the differences in performance are small and the order of the three
models mirror the prior probability they place on the independent-divergences
model.

We see similar, albeit more subtle, patterns when we look at the estimated
number of divergence events
(\figs S\ref{fig:neventsgrid} and S\ref{fig:neventsfixedcomparison}).

\begin{figure}[htbp]
    \begin{center}
        \includegraphics[width=\textwidth,height=\textheight,keepaspectratio]{../images/from-project-repo/model-performance-violin-cropped.pdf}
        \captionsetup{listformat=figList}
        \caption{
        Comparing the models' ability to estimate the divergence model.
        Each column of plots compares the performance of the Dirichlet process
        (DP), Pitman-Yor process (PYP) and weighted-uniform (WU) models when
        analyzing 720 \datasets simulated under the (from left-to-right)
        simultaneous-divergence, DP, PYP, WU, and independent-divergences
        model. 
        Mean \etimesets error is the mean partition distance
        of the posterior samples from the true
        divergence model (\etimesets),
        $p(\etimesets \given \alldata)$ is the posterior probability
        of the true divergence model,
        $p(\etimesets \given \alldata)$ is the posterior probability of the
        true divergence model,
        and
        $p(\etimesets \in \textrm{95\% CS})$ is the proportion of simulation
        replicates for which he true \etimesets is included in the 95\%
        posterior credibility set of divergence models.
        The bars overlaying the violin plots show the interval between the
        2.5\% and 97.5\% percentiles, and the white dot is the median.
        We generated the plot using matplotlib Version 3.1.3
        \citep{matplotlib}.
        }
        \label{fig:modelperformancegrid}
    \end{center}
\end{figure}

\begin{figure}[htbp]
    \begin{center}
        \includegraphics[width=\textwidth,height=\textheight,keepaspectratio]{../images/from-project-repo/nchars-model-performance-violin-cropped.pdf}
        \captionsetup{listformat=figList}
        \caption{
        Comparing the models' ability to estimate the independent-divergence
        model as the \dataset size increases.
        Each column of plots compares the performance of the Dirichlet process
        (DP), Pitman-Yor process (PYP) and weighted-uniform (WU) models when
        analyzing 720 \datasets with 5,000 to 500,000 characters simulated
        under the independent-divergences model
        Mean \etimesets error is the mean partition distance
        of the posterior samples from the true
        divergence model (\etimesets),
        $p(\etimesets \given \alldata)$ is the posterior probability
        of the true divergence model,
        $p(\etimesets \given \alldata)$ is the posterior probability of the
        true divergence model,
        and
        $p(\etimesets \in \textrm{95\% CS})$ is the proportion of simulation
        replicates for which he true \etimesets is included in the 95\%
        posterior credibility set of divergence models.
        The bars overlaying the violin plots show the interval between the
        2.5\% and 97.5\% percentiles, and the white dot is the median.
        We generated the plot using matplotlib Version 3.1.3
        \citep{matplotlib}.
        }
        \label{fig:modelperformancebysize}
    \end{center}
\end{figure}
